Improvement of Deterministic NVT Molecular Dynamics via Backward Error Analysis

### 2018-05-07 rigid_bodies.tex
corrected factor in omega
Separation of sentences for better diff tracking
Changes in citations in the introduction section

Molecular Simulation, being the computational realization of Statistical Mechanics,\cite{Tuckerman_2010} involves several types of approximations in order to mimic ``real'' physical systems. Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field. However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behaviour. For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Silveira_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD. That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.

The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules.\cite{Knorowski_2012, Patra_2013} As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj_2005} which is implemented, for instance, in the softwares packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS,\cite{Plimpton_1995} has been applied to simulate small molecules,\cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2} as well as membranes,\cite{Bucior_2012} molecular motors,\cite{Akimov_2012} micelles\cite{Yan_2008} and nanoparticles.\cite{Patra_2014}

The Nos\'{e}-Hoover chain thermostat,\cite{Martyna_1992} routinely used in MD, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables. The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with which backward error analysis is no longer feasible. Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi_2001, Ezra_2004, Ezra_2006}

In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body MD in the canonical (NVT) ensemble. The starting point is our previous work on the microcanonical (NVE) ensemble,\citep{Silveira_2017} which employs the unit quaternion representation for the rotational degrees of freedom. Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value. The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj_2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane_1987} which is exact in the sense that truncation errors do not affect its accuracy. We also include results from simulations using the conventional SHAKE\cite{Ryckaert_1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.

@@

Overall:
\begin{equation*}
\Ham S = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \tilde{\mt \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right] + \tilde{U} + \mathcal{O}({\timestep}^4),
\end{equation*}
where
\begin{equation*}
\tilde{U} = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right)
\end{equation*}
and
\begin{equation*}
\tilde{\mt \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} + {\mt \Gamma} \right)
\end{equation*}
with
\begin{equation*}
\mt \Omega = \left[\begin{array}{cc}
{\mt M}^{-1} & \mt 0 \\
\mt 0 & \frac{1}{2} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
\end{array}\right],
\end{equation*}
\begin{equation*}
\mt \Gamma = \left[\begin{array}{cc}
\mt 0 & \mt 0 \\
\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} {\mt S}({\mt I}^{-1} {\vt \tau}_b) \tr{\mt B}(\vt q)
\end{array}\right],
\end{equation*}
and
\begin{equation*}
{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
U_{\vt r \vt r} & U_{\vt r \vt q} \\
U_{\vt q \vt r} & U_{\vt q \vt q}
\end{array}\right].
\end{equation*}

@@

%\begin{center}
%  \label{fig:pressure}
%  \includegraphics{pressure2}
%  \captionof{figure}{Histograms of pressure obtained in this work (filled bars) and using LAMMPS %(unfilled bars). Simulation of liquid water with NVE integration. $\timestep : 1 fs$  }
%\end{center}

\section{Concluding Remarks}
\label{sec:conclusion}
Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$. 
by contrast

>>>>

Molecular Simulation, being the computational realization of Statistical Mechanics \cite{Tuckerman_2010}, involves several types of approximations in order to mimic ``real'' physical systems.
Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field.
However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behavior.
For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property \cite{Eastwood_2010}.
In a previous paper \cite{Silveira_2017}, we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD.
That finding brings into question the method introduced by Kamberaj and co-workers \cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.

The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules \cite{Knorowski_2012, Patra_2013}.
As a matter of fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented, for instance, in the softwares packages HOOMD-\textit{blue} \cite{Anderson_2008} and LAMMPS \cite{Plimpton_1995}, has been applied to simulate small molecules,\cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2} as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008} and nanoparticles \cite{Patra_2014}.

The Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992}, routinely used in MD, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables.
The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with which backward error analysis is no longer feasible.
Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers \cite{Sergi_2001, Ezra_2004, Ezra_2006}.

In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body MD in the canonical (NVT) ensemble.
The starting point is our previous work on the microcanonical (NVE) ensemble \cite{Silveira_2017}, which employs the unit quaternion representation for the rotational degrees of freedom.
Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value.
The analysis of the integrators includes comparisons with results from the method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane_1987} which is exact in the sense that truncation errors do not affect its accuracy.
We also include results from simulations using the conventional SHAKE \cite{Ryckaert_1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.

@@

Overall:
\begin{equation*}
\Ham S = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \tilde{\mt \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right] + \tilde{U} + \mathcal{O}({\timestep}^4),
\end{equation*}
where
\begin{equation*}
\tilde{U} = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right)
\end{equation*}
and
\begin{equation*}
\tilde{\mt \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} + {\mt \Gamma} \right)
\end{equation*}
with
\begin{equation*}
\mt \Omega = \left[\begin{array}{cc}
{\mt M}^{-1} & \mt 0 \\
\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
\end{array}\right],
\end{equation*}
\begin{equation*}
\mt \Gamma = \left[\begin{array}{cc}
\mt 0 & \mt 0 \\
\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} {\mt S}({\mt I}^{-1} {\vt \tau}_b) \tr{\mt B}(\vt q)
\end{array}\right],
\end{equation*}
and
\begin{equation*}
{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
U_{\vt r \vt r} & U_{\vt r \vt q} \\
U_{\vt q \vt r} & U_{\vt q \vt q}
\end{array}\right].
\end{equation*}

@@

%\begin{center}
%  \label{fig:pressure}
%  \includegraphics{pressure2}
%  \captionof{figure}{Histograms of pressure obtained in this work (filled bars) and using LAMMPS %(unfilled bars).
%Simulation of liquid water with NVE integration.
%$\timestep : 1 fs$  }
%\end{center}

\section{Concluding Remarks}
\label{sec:conclusion}
Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$.

by contrast
